A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-08-26 text/manuscript.Rmd
Add supplemental file text ref and hyperlink

We investigated this dataset further using multiple correspondence analysis
(MCA)  [@JSSv025i01], and visually summarized the analysis with UMAP (**figure
2**) [@McInnes2018]. We used the resulting UMAP graph coordinates to perform
density based clustering with HDBSCAN [@hdbscan], which resulted in
identification of 10 well-defined clusters capturing about 80% of the tumor
samples. To capture the remaining samples into one of these 10 clusters we used
kNN (see **Supplementary Methods** for details on clustering methods).

@@

Having defined tumors in terms of their pathway disruption profile, we sought to
understand whether different cancer types segregated into one or more
predominant classes. To our surprise, most cancer types were not heavily biased
in one class, and all well-represented cancer types had tumors in every class
(see **figure 3A** and full tumor profiles in **supplementary figure S1**, see
also interactive media from **supplementary file 1**), suggesting that,
in principle at least, our pathway-disruptions identify clusters of molecular
pathology largely independent of tissue-of-origin. As an example of one type of
cancer that does have a biased pathway profile, pancreatic adenocarcinoma (PAAD)
was predominantly found in class 8 (**figure 3A** and **supplementary figure
S1**). But even PAAD comprises tumors from the nine remaining classes, meaning
that patients suffering these tumors have potentially different underlying
molecular pathologies.

@@

Many cancers have molecular or histological subtypes defined based on gene
expression or other -omics profiles or pathology lab results. These subtypes
often have different standards of care owing to different overall drug
sensitivity (or other factors). If the histological subtypes represent true
molecular phenotypes, one predicts that histological subtypes should segregate
with our pathway-based clusters, therefore providing support for the clusters as
proxies for molecular pathology sub-typing. To our surprise, we find a similar
result to the previous analysis of cancer types projected onto the UMAP of
pathway disruptions. To illustrate this, we projected annotations for each of
the breast cancer subtypes, composed of Triple-negative/Basal-like, Her2
positive, normal-like, and luminal A and B subtypes onto the UMAP. These are
among the most heavily studied molecular subtypes in cancer, which each have
different prognoses and standards of care. We did not observe any exclusive
segregation by pathway for these subtype annotations (**Figure 3B** and
**supplemental file 1**). We also projected histological subtype data for the
remaining cancers (see **supplementary figure S2** for the full set of
projections); we find that the subtypes, though often biased towards one or more
classes, are almost never exclusive. We interpret these data in aggregate to
mean that our pathway disruption classes do not correspond to previously
identified molecular subtypes within the parent cancer type.

@@

There are several well-known familial cancer-causing mutations that have been
interrogated extensively for differences in basic biology, survival and
treatment outcomes. However, the functions of these genes are related to risk
factors such as genome stability generally, proof-reading and DNA damage repair,
and telomere length. _BRCA1/2_ genes for example are key for DNA double-stranded
break repair [@moynihan-1999-brca1;@davies-2001-brca2] and germline mutations in
these genes confer elevated risk for breast, prostate and ovarian cancers. The
mechanism of risk is thought to involve loss of heterozygosity resulting in loss
of the wildtype, functional allele [@brca-loh], so we projected the somatic
mutations for _BRCA1_ and _BRCA2_ genes onto the UMAP, but did not observe
segregation of these mutations into specific clusters (**figure 3C**). We made
similar projections for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH1_,
_MLH3_, _PMS1_ and _PMS2_; _BRIP1_, _RAD51_, _CHEK2_ and _APC_. None of these
genes except for _APC_ exhibited any remarkable specificity with respect to
cluster assignment (**supplementary figure S3**). To look at other risk factors
such as maintenance of DNA methylation levels and telomere length, we projected
somatic mutations of the _TET2_ and _TET3_ genes, plus _TERT_, _TEP1_, and
_DKC1_, and observed similar lack of segregation by cluster (**supplementary
figure S3**).

@@

To discover what pathways are most important for clustering, we calculated
percent enrichment for each pathway _within cluster_ relative to all other
clusters combined (see methods) and ranked pathways for each cluster from
highest to lowest enrichment. We visualized the enrichment as a heatmap
(**figure 5A**). Using this approach, we identified 25 pathways highly enriched
(enrichment score ≥ 0.3, 95% confidence; see methods) in cluster 2, 9 pathways
enriched in cluster 3, 24 pathways in cluster 4, 10 pathways in cluster 5, 10
pathways in cluster 6, 24 pathways in cluster 7, 24 pathways in cluster 8, 10
pathways in cluster 9, and 0 pathways for clusters 1 and 10 (**supplementary table ST2**). Clusters 7, 8, and
9 had several pathways in common. To explore the specific pathways marking each
cluster, we projected disruptions for each of the 377 pathways onto the UMAP
(**supplementary figure S4**). Clusters 3 and 5 were distinguished by metabolic
pathways including RNA and protein biosynthesis (**supplementary figure S4**).
Similarly, cluster 4 was distinguished by mutations affecting regulation of DNA
and histone methylation ("DNA methylation", "PRC2 methylates histones and DNA",
and "Nucleosome assembly"). Clusters 7-9 have in common mutations in
extracellular, intracellular, and immune-related signaling pathways (see
**figure 5B** and **supplementary figure S4**). Cluster 2 had the highest
pathway enrichment levels of the three, having mutations in hedgehog signaling,
"β-catenin degradation", "cellular response to hypoxia", "regulation of cell
cycle" and "apoptosis" among others.

@@

Prior efforts to extract common signatures from pan-cancer datasets met with
difficulty in distinguishing tumor samples from their tissue-specific -omics
data signatures. Given our pathway-disruption based clustering, this raises
the question, are tumor phenotypes driven entirely by common driver genes, or by
“silent” tissue-specific effectors ( _i.e._ too few samples to detect above
statistical significance thresholds), or a combination of both? To answer this
question, we compared top pathway genes for each cluster relative to the TCGA
background to find differentially mutated genes. We ranked odds ratios and
selected the top ten enriched and depleted genes (pvalue < 0.01) for each
cluster (**Figure 6**; odds ratios plot). Clusters 7 and 8, which shared
multiple enrichment in signaling pathways, are largely driven by mutations in
_PI3K_ and its orthologs and _Ras_ genes, respectively (compare _PIK3CA_ and
_KRAS_ panels of **Supplementary figure S5**). Interestingly, cluster 9, which
also shared multiple enrichment in signaling pathways with clusters 7 and 8, is
enriched for both _PIK3CA_ and _KRAS_. Clusters 3 and 5, defined by enrichment
in metabolic pathways, had mutations in ribosomal proteins and nuclear pore
complex, respectively. Cluster 4 had mutations in genes responsible for
nucleosome structure. Cluster 2 had mutations in proteasomal subunit genes
involved in protein degradation. We also observed that genes that were enriched
for one cluster are depleted from others ( _e.g. TP53_ is enriched in cluster 6,
but depleted in cluster 7; _PIK3CA_ is enriched in cluster 7, but depleted in
clusters 3 and 8). Next, we investigated the proportion of samples per cancer
type for the significant genes within a cluster (**Figure 6**; heatmap).
Surprisingly, cluster-specific tumors were not predominated by one or more
highly mutated genes across all cancers. Instead, when observing the mutation
rate for these genes within samples that belong to a cluster, the mutation rate
is heterogeneous across tumors by tissue origin ( _e.g._ In cluster 4, CESC was
enriched for _H2AFX_, OV was enriched for _HIST1H2BD_, and UCEC was enriched for
_HIST1H2AC_). Even among the top most mutationally enriched genes within
clusters there is no global pattern, indicating that our clusters are not driven
by individual genes, but rather networks as a whole. Taken together, our data
identifies a framework of cancer type-specific mutations associated with
specific clusters.

>>>>

We investigated this dataset further using multiple correspondence analysis
(MCA)  [@JSSv025i01], and visually summarized the analysis with UMAP (**figure
2** and see interactive media from [**supplemental file 1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)) [@McInnes2018]. 
We used the resulting UMAP graph coordinates to perform
density based clustering with HDBSCAN [@hdbscan], which resulted in
identification of 10 well-defined clusters capturing about 80% of the tumor
samples. To capture the remaining samples into one of these 10 clusters we used
kNN (see **Supplementary Methods** for details on clustering methods).

@@

Having defined tumors in terms of their pathway disruption profile, we sought to
understand whether different cancer types segregated into one or more
predominant classes. To our surprise, most cancer types were not heavily biased
in one class, and all well-represented cancer types had tumors in every class
(see **figure 3A** and full tumor profiles in **supplementary figure S1**, see
also interactive media from **supplementary file 1**), suggesting that,
in principle at least, our pathway-disruptions identify clusters of molecular
pathology largely independent of tissue-of-origin. As an example of one type of
cancer that does have a biased pathway profile, pancreatic adenocarcinoma (PAAD)
was predominantly found in class 8 (**figure 3A** and **supplementary figure
S1** and **supplementary file 1**). But even PAAD comprises tumors from the nine remaining classes, meaning
that patients suffering these tumors have potentially different underlying
molecular pathologies.

@@

Many cancers have molecular or histological subtypes defined based on gene
expression or other -omics profiles or pathology lab results. These subtypes
often have different standards of care owing to different overall drug
sensitivity (or other factors). If the histological subtypes represent true
molecular phenotypes, one predicts that histological subtypes should segregate
with our pathway-based clusters, therefore providing support for the clusters as
proxies for molecular pathology sub-typing. To our surprise, we find a similar
result to the previous analysis of cancer types projected onto the UMAP of
pathway disruptions. To illustrate this, we projected annotations for each of
the breast cancer subtypes, composed of Triple-negative/Basal-like, Her2
positive, normal-like, and luminal A and B subtypes onto the UMAP. These are
among the most heavily studied molecular subtypes in cancer, which each have
different prognoses and standards of care. We did not observe any exclusive
segregation by pathway for these subtype annotations (**Figure 3B**). We also 
projected histological subtype data for the remaining cancers (see 
**supplementary figure S2** for the full set of projections and see interactive
media from [**supplemental file 1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)); we find that the subtypes, though often 
biased towards one or more classes, are almost never exclusive. We interpret 
these data in aggregate to mean that our pathway disruption classes do not 
correspond to previously identified molecular subtypes within the parent cancer type.

@@

There are several well-known familial cancer-causing mutations that have been
interrogated extensively for differences in basic biology, survival and
treatment outcomes. However, the functions of these genes are related to risk
factors such as genome stability generally, proof-reading and DNA damage repair,
and telomere length. _BRCA1/2_ genes for example are key for DNA double-stranded
break repair [@moynihan-1999-brca1;@davies-2001-brca2] and germline mutations in
these genes confer elevated risk for breast, prostate and ovarian cancers. The
mechanism of risk is thought to involve loss of heterozygosity resulting in loss
of the wildtype, functional allele [@brca-loh], so we projected the somatic
mutations for _BRCA1_ and _BRCA2_ genes onto the UMAP, but did not observe
segregation of these mutations into specific clusters (**figure 3C** and see 
interactive media from [**supplemental file 1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). We made similar projections 
for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH1_, _MLH3_, _PMS1_ 
and _PMS2_; _BRIP1_, _RAD51_, _CHEK2_ and _APC_. None of these genes except 
for _APC_ exhibited any remarkable specificity with respect to cluster assignment 
(**supplementary figure S3**). To look at other risk factors
such as maintenance of DNA methylation levels and telomere length, we projected
somatic mutations of the _TET2_ and _TET3_ genes, plus _TERT_, _TEP1_, and
_DKC1_, and observed similar lack of segregation by cluster (**supplementary
figure S3**).

@@

To discover what pathways are most important for clustering, we calculated
percent enrichment for each pathway _within cluster_ relative to all other
clusters combined (see methods) and ranked pathways for each cluster from
highest to lowest enrichment. We visualized the enrichment as a heatmap
(**figure 5A**). Using this approach, we identified 25 pathways highly enriched
(enrichment score ≥ 0.3, 95% confidence; see methods) in cluster 2, 9 pathways
enriched in cluster 3, 24 pathways in cluster 4, 10 pathways in cluster 5, 10
pathways in cluster 6, 24 pathways in cluster 7, 24 pathways in cluster 8, 10
pathways in cluster 9, and 0 pathways for clusters 1 and 10 (**supplementary table ST2**). Clusters 7, 8, and
9 had several pathways in common. To explore the specific pathways marking each
cluster, we projected disruptions for each of the 377 pathways onto the UMAP
(**supplementary figure S4** and see interactive media from [**supplemental file 1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). 
Clusters 3 and 5 were distinguished by metabolic pathways including RNA and 
protein biosynthesis (**supplementary figure S4**). Similarly, cluster 4 was 
distinguished by mutations affecting regulation of DNA
and histone methylation ("DNA methylation", "PRC2 methylates histones and DNA",
and "Nucleosome assembly"). Clusters 7-9 have in common mutations in
extracellular, intracellular, and immune-related signaling pathways (see
**figure 5B** and **supplementary figure S4**). Cluster 2 had the highest
pathway enrichment levels of the three, having mutations in hedgehog signaling,
"β-catenin degradation", "cellular response to hypoxia", "regulation of cell
cycle" and "apoptosis" among others.

@@

Prior efforts to extract common signatures from pan-cancer datasets met with
difficulty in distinguishing tumor samples from their tissue-specific -omics
data signatures. Given our pathway-disruption based clustering, this raises
the question, are tumor phenotypes driven entirely by common driver genes, or by
“silent” tissue-specific effectors ( _i.e._ too few samples to detect above
statistical significance thresholds), or a combination of both? To answer this
question, we compared top pathway genes for each cluster relative to the TCGA
background to find differentially mutated genes. We ranked odds ratios and
selected the top ten enriched and depleted genes (pvalue < 0.01) for each
cluster (**Figure 6**; odds ratios plot). Clusters 7 and 8, which shared
multiple enrichment in signaling pathways, are largely driven by mutations in
_PI3K_ and its orthologs and _Ras_ genes, respectively (compare _PIK3CA_ and
_KRAS_ panels of **Supplementary figure S5** and see interactive media from 
[**supplemental file 1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). Interestingly, cluster 9, which
also shared multiple enrichment in signaling pathways with clusters 7 and 8, is
enriched for both _PIK3CA_ and _KRAS_. Clusters 3 and 5, defined by enrichment
in metabolic pathways, had mutations in ribosomal proteins and nuclear pore
complex, respectively. Cluster 4 had mutations in genes responsible for
nucleosome structure. Cluster 2 had mutations in proteasomal subunit genes
involved in protein degradation. We also observed that genes that were enriched
for one cluster are depleted from others ( _e.g. TP53_ is enriched in cluster 6,
but depleted in cluster 7; _PIK3CA_ is enriched in cluster 7, but depleted in
clusters 3 and 8). Next, we investigated the proportion of samples per cancer
type for the significant genes within a cluster (**Figure 6**; heatmap).
Surprisingly, cluster-specific tumors were not predominated by one or more
highly mutated genes across all cancers. Instead, when observing the mutation
rate for these genes within samples that belong to a cluster, the mutation rate
is heterogeneous across tumors by tissue origin ( _e.g._ In cluster 4, CESC was
enriched for _H2AFX_, OV was enriched for _HIST1H2BD_, and UCEC was enriched for
_HIST1H2AC_). Even among the top most mutationally enriched genes within
clusters there is no global pattern, indicating that our clusters are not driven
by individual genes, but rather networks as a whole. Taken together, our data
identifies a framework of cancer type-specific mutations associated with
specific clusters.
